---
title: "Weight Lifting Exercise Prediction"
subtitle: "Practicle Machine Learning Course Project"
author: "Brian Gulbis"
date: "August 16, 2015"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

```{r}
library(dplyr)
library(lubridate)
library(caret)

library(doParallel)
if (Sys.getenv("PROCESSOR_ARCHITECTURE") != "x86") {
    registerDoParallel(cores=4)
}
```


```{r warning=FALSE}
training <- read.csv("pml-training.csv", stringsAsFactors = FALSE) %>%
    mutate(cvtd_timestamp = dmy_hm(cvtd_timestamp),
           user_name = factor(user_name),
           new_window = factor(new_window),
           classe = factor(classe)) %>%
    rename(max_pitch_belt = max_picth_belt,
           kurtosis_pitch_belt = kurtosis_picth_belt,
           kurtosis_pitch_arm = kurtosis_picth_arm,
           max_pitch_arm = max_picth_arm,
           max_pitch_dumbbell = max_picth_dumbbell,
           kurtosis_pitch_forearm = kurtosis_picth_forearm,
           max_pitch_forearm = max_picth_forearm) %>%
    mutate_each(funs(as.numeric), roll_belt:magnet_forearm_z)

testing <- read.csv("pml-testing.csv", stringsAsFactors = FALSE) %>%
    mutate(cvtd_timestamp = dmy_hm(cvtd_timestamp),
           user_name = factor(user_name),
           new_window = factor(new_window)) %>%
    rename(max_pitch_belt = max_picth_belt,
           kurtosis_pitch_belt = kurtosis_picth_belt,
           kurtosis_pitch_arm = kurtosis_picth_arm,
           max_pitch_arm = max_picth_arm,
           max_pitch_dumbbell = max_picth_dumbbell,
           kurtosis_pitch_forearm = kurtosis_picth_forearm,
           max_pitch_forearm = max_picth_forearm) %>%
    mutate_each(funs(as.numeric), roll_belt:magnet_forearm_z)
    
```

Exploratory Analysis

```{r}
summary(training)

featurePlot(x=training[,c("avg_roll_belt","var_roll_belt")], y=training$classe, plot="pairs")

qplot(var_roll_belt, classe, data=training)

```

Potential features to include (from authors)

* Belt
  * mean and variance of roll: avg\_roll\_belt, var\_roll\_belt
  * maximum, range, and variance of accelerometer: var\_total\_accel\_belt (?)
  * variance of gyro
  * variance of magnetometer
* Arm
  * variance of accelerometer
  * maximum and minimum of magnetometer
* Dumbbell
  * maximum of accelerometer
  * variance of gyro
  * maximum and minimum of magnetometer
* Forearm
  * sum of pitch: pitch\_forearm
  * maximum and minimum of gyro

```{r}


```


Near Zero Frequency

```{r}
#folds <- createFolds(y=training$classe)

#near.zero <- nearZeroVar(training, saveMetrics = TRUE)
trainFtr <- select(training, -(X:new_window), -classe)
nzv <- nearZeroVar(trainFtr)
fltr.nz <- trainFtr[, -nzv]

```

Correlated Predictors

```{r}
corPred <- cor(fltr.nz, use="complete.obs")
highCor <- findCorrelation(corPred, cutoff = 0.8)
fltr.cor <- fltr.nz[, -highCor]
#highCor <- abs(cor.pred[upper.tri(cor.pred)]) > 0.9
#sum(highCor)
#summary(cor.pred[upper.tri(cor.pred)])

```

Center and Scale; Impute

```{r}
set.seed(1235)
#preProc <- preProcess(fltr.cor, method=c("center", "scale"))
preProc <- preProcess(fltr.cor, method=c("knnImpute"))
train.preproc <- predict(preProc, fltr.cor)
train.preproc$classe <- training$classe
```


```{r}
modelGLM <- train(classe ~ ., method="glm", data=train.preproc)
```

```{r}
library(rpart)
modelTree <- train(classe ~ ., method="rpart", data=, trControl=trainControl("repeatedcv"))
predTree <- predict(modelTree, testing[,near.zero$zeroVar == FALSE])
```

```{r}
modelRF <- train(classe ~ ., method="rf", data=train.preproc, trControl=trainControl("repeatedcv"))
```


```{r warning=FALSE}
near.zero <- nearZeroVar(training, saveMetrics = TRUE)
nz <- training[,near.zero$zeroVar == FALSE]
preProc <- preProcess(nz[,8:150], method="pca", pcaComp=5)
trainPCA <- predict(preProc, nz[,8:150])
modelFit <- train(training$classe ~ ., method="glm", data=trainPCA)
```

