---
title: "Weight Lifting Exercise Prediction"
subtitle: "Practicle Machine Learning Course Project"
author: "Brian Gulbis"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

```{r echo=FALSE}
library(plyr)
library(dplyr)
library(lubridate)
library(caret)
```

```{r echo=FALSE}
library(doParallel)
if (Sys.getenv("PROCESSOR_ARCHITECTURE") != "x86") {
    registerDoParallel(core=4)
}
```

```{r warning=FALSE}
training <- read.csv("pml-training.csv", stringsAsFactors = FALSE) %>%
    mutate(cvtd_timestamp = dmy_hm(cvtd_timestamp),
           user_name = factor(user_name),
           new_window = factor(new_window),
           classe = factor(classe)) %>%
    rename(max_pitch_belt = max_picth_belt,
           kurtosis_pitch_belt = kurtosis_picth_belt,
           kurtosis_pitch_arm = kurtosis_picth_arm,
           max_pitch_arm = max_picth_arm,
           max_pitch_dumbbell = max_picth_dumbbell,
           kurtosis_pitch_forearm = kurtosis_picth_forearm,
           max_pitch_forearm = max_picth_forearm) %>%
    mutate_each(funs(as.numeric), roll_belt:magnet_forearm_z)

testing <- read.csv("pml-testing.csv", stringsAsFactors = FALSE) %>%
    mutate(cvtd_timestamp = dmy_hm(cvtd_timestamp),
           user_name = factor(user_name),
           new_window = factor(new_window)) %>%
    rename(max_pitch_belt = max_picth_belt,
           kurtosis_pitch_belt = kurtosis_picth_belt,
           kurtosis_pitch_arm = kurtosis_picth_arm,
           max_pitch_arm = max_picth_arm,
           max_pitch_dumbbell = max_picth_dumbbell,
           kurtosis_pitch_forearm = kurtosis_picth_forearm,
           max_pitch_forearm = max_picth_forearm) %>%
    mutate_each(funs(as.numeric), roll_belt:magnet_forearm_z)
```

## Create Validation Partition

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.75, list=FALSE)
train.data <- training[inTrain,]
valid.data <- training[-inTrain,]
```

## Exploratory Analysis

```{r}
#summary(training)

#featurePlot(x=training[,c("avg_roll_belt","var_roll_belt")], y=training$classe, plot="pairs")

#qplot(var_roll_belt, classe, data=training)

```

Potential features to include (from authors)

* Belt
  * mean and variance of roll: avg\_roll\_belt, var\_roll\_belt
  * maximum, range, and variance of accelerometer: var\_total\_accel\_belt (?)
  * variance of gyro
  * variance of magnetometer
* Arm
  * variance of accelerometer
  * maximum and minimum of magnetometer
* Dumbbell
  * maximum of accelerometer
  * variance of gyro
  * maximum and minimum of magnetometer
* Forearm
  * sum of pitch: pitch\_forearm
  * maximum and minimum of gyro


## Near Zero Frequency

Remove near-zero variables

```{r}
train.set <- select(train.data, -(X:new_window), -classe)
near.zero <- nearZeroVar(train.set)
train.set <- train.set[, -near.zero]
```

## Find Highly-Correlated Predictors

Find highly-correlated predictors and remove

```{r}
train.cor <- cor(train.set, use="complete.obs")
high.cor <- findCorrelation(train.cor, cutoff = 0.8)
train.set <- train.set[, -high.cor]
```

## Center and Scale; Impute Missing Values

Use k-nearest neighbor to impute missing values

```{r}
set.seed(1235)
pre.proc <- preProcess(train.set, method=c("knnImpute"))
train.set <- predict(pre.proc, train.set)
train.set$classe <- train.data$classe
```

## Prep Training Validation Set

```{r}
valid.set <- select(valid.data, -(X:new_window), -classe)
valid.set <- valid.set[, -near.zero]
valid.set <- valid.set[, -high.cor]
valid.set <- predict(pre.proc, valid.set)
valid.set$classe <- valid.data$classe
```

## Prep Final Test Set

```{r}
test.set <- dplyr::select(testing, -(X:new_window), -problem_id)
test.set <- test.set[, -near.zero]
test.set <- test.set[, -high.cor]
test.set <- predict(pre.proc, test.set)
```


```{r echo=FALSE}
myseeds <- list(c(123,234,345), c(456,567,678), c(789,890,901), c(987,876,765), c(654,543,432), c(321,210,109), c(135,246,357), c(468,579,680), c(791,802,913), c(975,864,753), 54321) 
```

## Prediction Models

Use Random Forest to create predication model

```{r cache=TRUE}
trCtrl <- trainControl(method="repeatedcv", seeds=myseeds)

modelRF.save <- "modelRF.Rds"

if (file.exists(modelRF.save)) {
    # Read the model in and assign it to a variable.
    library(randomForest)
    modelRF <- readRDS(modelRF.save)
} else {
    # Otherwise, run the training.
    modelRF <- train(classe ~ ., method="rf", data=train.set, trControl=trCtrl)
    saveRDS(modelRF, modelRF.save)
}
```

Boosting

```{r}
modelGBM.save <- "modelGBM.Rds"

if (file.exists(modelGBM.save)) {
    # Read the model in and assign it to a variable.
    library(gbm)
    modelGBM <- readRDS(modelGBM.save)
} else {
    # Otherwise, run the training.
    modelGBM <- train(classe ~ ., method="gbm", data=train.set, verbose=FALSE, trControl=trCtrl)
    saveRDS(modelGBM, modelGBM.save)
}
```

LDA

```{r}
modelLDA <- train(classe ~ ., method="lda", data=train.set, trControl=trCtrl)
```

## Compare Prediction Models

```{r}
predRF <- predict(modelRF, valid.set)
cmRF <- confusionMatrix(predRF, valid.set$classe)

predGBM <- predict(modelGBM, valid.set)
cmGBM <- confusionMatrix(predGBM, valid.set$classe)

predLDA <- predict(modelLDA, valid.set)
cmLDA <- confusionMatrix(predLDA, valid.set$classe)
```

## Create a Stacked Model

```{r}
predDF <- data.frame(predRF, predGBM, classe=valid.set$classe)
mod.comb <- train(classe ~ ., data=predDF, method="rf")
pred.comb <- predict(mod.comb, valid.set)
confusionMatrix(pred.comb, valid.set$classe)

```

## Predict Testing Set

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
```

```{r}
predTest <- predict(modelRF, test.set)
answer <- as.character(predTest)
pml_write_files(answer)
saveRDS(predTest, "test_predictions.Rds")
```

